Machine Learning can Advance Theory Formation in the Social Sciences

### 2023-07-12 presentation_isdp2023.qmd
update 2023-07-12 12:23:25.619832

---
title: "Machine Learning can Advance Theory Formation in the Social Sciences"
subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>[doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
author: "Caspar J. Van Lissa"
institute: "Tilburg University, dept. Methodology & Statistics"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
format: revealjs
bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
---

@@

* Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
* Current solutions improve deductive research (theory-testing)
    + questionable research practices [@wichertsDegreesFreedomPlanning2016]
    + replication
    + preregistration

@@

**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.

* Most social scientific theories are:
    + insufficiently precise
    + too flexible
* Compare to @walasekEvolutionaryModelSensitive2022

@@

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

* Confirmatory methods poorly suited to exploratory research
    + p-values: Type I error is 5%, but false discovery rate (FDR) is 14-50% [@vidgenPValuesMisunderstoodMisused2016].
    + Model fit indices: manually specify models, no guarantee that best model is included
* Role of flexibility: there are infinite ways to explore
* "unguided exploration" is effortful and inflates the risk of spurious results

@@

* automated model building
* checks and balances to prevent overfitting
* maximize predictive performance and generalizability of results

@@

* Phenomena: stable and general features of the world
* Few quantitative methods for phenomena detection (aside from expert opinion)

**Text mining** may be suitable for phenomena detection [@vanlissaMappingPhenomenaRelevant2021]

* Unsupervised learning method
* Clusters in keywords and abstracts

@@

* Good theory incorporates most important causes
* Cannot assess the relative utility of predictors across studies
* Include potential predictors from various theories, and undertheorized factors
* Pitfalls:
    + Multicollinearity
    + Colliders

@@

* Machine learning can provide nuance by revealing them

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

* E.g., latent class analyses
    + "Which individuals are similar?"
    + Other *unsupervised learning* methods exist
* Tree-based models
    + "Why are these individuals similar?"
    + Group individuals based on predictors to maximize homogeneity of the outcome
* RI-CLPM and DSEM
    + Rarely explain heterogeneity in within-person effects
    + Regularization

@@

Checks and balances to curtail *overfitting*

* cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]

@@

* Variable importance metrics
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?

* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?

Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model

## From models to theory 2

Data models -> formal theory [@haslbeckModelingPsychopathologyData2021]

* Abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

## Text Mining Systematic Review

* Identify which constructs are discussed in 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
* Identify which constructs co-occur in papers
* Compare resulting networks to construct networks from existing theories and systematic reviews
* Paper: <https://doi.org/10.1007/s40894-021-00160-7>
* Code and data: <https://github.com/cjvanlissa/veni_sysrev>
* Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]

## Baseline network {.smaller}

```{r}
#| label = "figbaseline",
#| warning = FALSE,
#| eval = TRUE,
#| results = "asis",
#| out.width = "100%",
#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
#| the theory)."
knitr::include_graphics("presentation/baseline_network.png") 
```

(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.

@@

* Both analyses reflected constructs from theoretical literature
    + Especially pertaining to neurodevelopment and socialization
* Mental health-related outcomes feature prominently
    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
    + This underlines the importance of research in this area
* Substantial correspondence between keywords and abstracts, suggesting validity of method
* Networks are sparse; few connections among constructs: fragmented literature

@@

<!-- ## Implications -->

<!-- * Empirical research relies on theory; under-theorized phenomena may be overlooked -->
<!--     + TMSR offers guidance for phenomena to consider as confounders/causes -->
<!-- * Unembedded terms indicate promising areas of future research -->
<!-- * TMSR network can serve as proto-theoretical nomological network -->

## Predicting ER development {.smaller}

* What are important predictors of adolescents' ER development?
* What is the nature of the association of the predictors with the trajectories?
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>

@@

@@

* Best predictors: Personality and conflict resolution behavior
* Negative parenting more predictive than other positive practices

@@

* Many predictors show non-linear effects

@@

* Machine learning reveals blind spots in existing theory
    + Missing constructs, relative importance, functional form
* ML can detect patterns in published research and primary data
* "Inductive methods" for consolidating insights into theory remain vague
    + Theories are still verbal and flexible, how should they be updated?
* Ignoring causality limits interpretability of machine learning
    + Incorrect direction of causality: meaningless results
    + Conditioning on a collider: biased results

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. https://doi.org/10.1002/icd.2370

Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior. https://doi.org/10.1080/2153599X.2022.2070254

Van Lissa, C. J., Beinhauer, L., Branje, S., & Meeus, W. (2022). Using machine learning to identify early predictors of adolescent emotion regulation development. Journal of Research on Adolescence. https://doi.org/10.31234/osf.io/rbqf9

Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. https://doi.org/10.1007/s40894-021-00160-7

Van Lissa, C. J., et al., (2022). Using machine learning to identify important predictors of COVID-19 infection prevention behaviors during the early phase of the pandemic. Patterns, 3(4), 100482. https://doi.org/10.1016/j.patter.2022.100482

>>>>

---
title: "Machine Learning can Advance Theory Formation in the Social Sciences"
subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>Paper: [doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
author: "Caspar J. Van Lissa"
institute: "Tilburg University, dept. Methodology & Statistics"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
format: revealjs
bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
---

@@

* Replication crisis [@scheelWhyMostPsychological2022; @lavelleWhenCrisisBecomes2021]
* Current solutions improve deductive research (theory-testing)
    + preregistration [@peikertReproducibleResearchTutorial2021]
    + replication
    + questionable research practices [@wichertsDegreesFreedomPlanning2016]

@@

@@

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers (parents, practitioners, policy makers)

@@

* Confirmatory methods poorly suited to exploratory research
    + p-values: 5% Type I error prob is consistent with 14-50% false discovery rate [@vidgenPValuesMisunderstoodMisused2016]
    + Depending on prior probability of hypotheses being true
    + Model fit indices: manually specify models, no guarantee that best model is in set
* Role of flexibility: there are infinite ways to explore
* "unguided exploration" is effortful and inflates the risk of spurious results

@@

* automated model building

@@

* Phenomena: stable features of the world reliably evidenced by patterns in data (Woodward, 1989)

Machine learning  [@vanlissaMappingPhenomenaRelevant2021]

* Unsupervised learning methods find patterns in data
* Text mining finds patterns in published literature

@@

* Good theory incorporates most important causes
* Cannot assess the relative utility of predictors across studies
* Include potential predictors from various theories, and undertheorized factors

@@

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* Regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

* Unsupervised learning methods
    + Group individuals to maximize homogeneity of some outcomes
    + "Which individuals are similar?"
* Tree-based models
    + Group individuals by predictors to maximize homogeneity of outcomes
    + "Why are these individuals similar?"
* Mixed-effects / multilevel models (DSEM, RI-CLPM)
    + Often model but rarely explain heterogeneity in within-person effects
    + Regularization (e.g., regularized SEM)

@@

Machine learning

* Incorporates checks and balances to curtail *overfitting*
* Maximizes predictive performance and generalizability of results
* Cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]
* Possible to estimate generalizability to representative subsample (e.g., Douglas, Sutton, Van Lissa et al., 2023)

@@

* Variable importance metrics
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?
* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?
* Abductive formal theory construction (AFTC) framework [@haslbeckModelingPsychopathologyData2021]
    + Translate insights into formal theory
    + Conduct "posterior predictive check" to check fit with data
    + Discrepancies: amend formal theory

@@

## Example: TMSR {.smaller}

* Identifying relevant phenomena
* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
* Text Mining Systematic Review [@vanlissaMappingPhenomenaRelevant2021]
    + Unlimited sample size, transparent, objective, reproducible
* 6653 papers on **emotion regulation** in **adolescence** [age 10-24]

Paper: <https://doi.org/10.1007/s40894-021-00160-7>
Reproducible code and data: <https://github.com/cjvanlissa/veni_sysrev>

Workflow for Open Reproducible Code in Science (WORCS)  [@vanlissaWORCSWorkflowOpen2020]

@@

* TM analyses retrieve constructs from theoretical literature
    + Especially pertaining to neurodevelopment and socialization
    + This suggests validity
* Substantial correspondence between keyword and abstract network
    + This suggests reliability
* Networks are sparse; few connections among constructs:
    + Fragmented literature, as noted by prior reviews

@@

## Implications

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + TMSR offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* TMSR network can serve as proto-theoretical nomological network

## Example: Predictors of Development {.smaller}

* What are important predictors of ER development trajectories?
* What is the nature of association of predictors with trajectories?
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>

@@

## Variable importance

```{r, out.width="90%"}
knitr::include_graphics("pcomb2.png")
```

## Variable importance 2

```{r, out.width="100%"}
knitr::include_graphics("violin.png")
```

## Marginal association

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp1-12.png")
```

@@

* Best predictors: Personality and conflict behaviors
* Negative parenting more predictive than positive parenting

@@

* Most predictors show non-linear effects

## Reflection content

* Proximal factors more predictive of ER development than distal predictors
    + Order mirrors bioecological model
* Most important predictors are routinely assessed (big 5) or overt (conflict)
  + Prime candidates for early risk assessment
  + Conflict resolution behavior can be taught
    - Target for intervention if association is causal
* Emphasis on parenting (in literature and theory) might not be justified
  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]

## Reflections form

* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
* Some predictors show almost no marginal effect (e.g., father's age, drug use)
  + This suggests they might be important in **interactions**

@@

* Machine learning reveals blind spots in existing theory
    + Missing constructs, relative importance, functional form
* ML can detect patterns in published research and primary data
* Does not solve the "inductive problem"
    + Still your responsibility to consolidate insights into theory
* Ignoring causality limits interpretability of machine learning
    + Incorrect direction of causality: meaningless results
    + Conditioning on a collider: biased results

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. ICD. DOI:10.1002/icd.2370

Van Lissa, C. J. (2022). Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning. Religion, Brain & Behavior. DOI:10.1080/2153599X.2022.2070254

Van Lissa, C. J., Beinhauer, L., Branje, S., & Meeus, W. (2022). Using machine learning to identify early predictors of adolescent emotion regulation development. Journal of Research on Adolescence. DOI:10.31234/osf.io/rbqf9

Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. ARR. DOI:10.1007/s40894-021-00160-7

Van Lissa, C. J., et al., (2022). Using machine learning to identify important predictors of COVID-19 infection prevention behaviors during the early phase of the pandemic. Patterns, 3(4), 100482. DOI:10.1016/j.patter.2022.100482

Douglas, K., Sutton, R. M., Van Lissa, C. J., Wolfgang, S., Jannis, K., Maximilian, A., ... & N Pontus, L. (2023). Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing. EJSP. DOI: 10.1002/ejsp.2968

### 2023-07-12 presentation_tuebingen2023.qmd
update 2023-07-12 12:23:25.619832

@@

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers (parents, practitioners, policy makers)

@@

@@

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* Regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

* Unsupervised learning methods
    + Group individuals to maximize homogeneity of some outcomes
    + "Which individuals are similar?"
* Tree-based models
    + Group individuals by predictors to maximize homogeneity of outcomes
    + "Why are these individuals similar?"
* Mixed-effects / multilevel models (DSEM, RI-CLPM)
    + Often model but rarely explain heterogeneity in within-person effects
    + Regularization (e.g., regularized SEM)

@@

Machine learning

@@

* Variable importance metrics
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?
* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?
* Abductive formal theory construction (AFTC) framework [@haslbeckModelingPsychopathologyData2021]
    + Translate insights into formal theory
    + Conduct "posterior predictive check" to check fit with data
    + Discrepancies: amend formal theory

@@

## Example: TMSR {.smaller}

* Identifying relevant phenomena
* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
* Text Mining Systematic Review [@vanlissaMappingPhenomenaRelevant2021]
    + Unlimited sample size, transparent, objective, reproducible
* 6653 papers on **emotion regulation** in **adolescence** [age 10-24]

Paper: <https://doi.org/10.1007/s40894-021-00160-7>
Reproducible code and data: <https://github.com/cjvanlissa/veni_sysrev>

Workflow for Open Reproducible Code in Science (WORCS)  [@vanlissaWORCSWorkflowOpen2020]

@@

* What are important predictors of ER development trajectories?
* What is the nature of association of predictors with trajectories?
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>

@@

@@

## Reflection content

* Proximal factors more predictive of ER development than distal predictors
    + Order mirrors bioecological model
* Most important predictors are routinely assessed (big 5) or overt (conflict)
  + Prime candidates for early risk assessment
  + Conflict resolution behavior can be taught
    - Target for intervention if association is causal
* Emphasis on parenting (in literature and theory) might not be justified
  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]

## Reflections form

* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
* Some predictors show almost no marginal effect (e.g., father's age, drug use)
  + This suggests they might be important in **interactions**

@@

* Machine learning reveals blind spots in existing theory
    + Missing constructs, relative importance, functional form
* ML can detect patterns in published research and primary data
* Does not solve the "inductive problem"
    + Still your responsibility to consolidate insights into theory
* Ignoring causality limits interpretability of machine learning
    + Incorrect direction of causality: meaningless results
    + Conditioning on a collider: biased results

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. ICD. DOI:10.1002/icd.2370

@@

Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. ARR. DOI:10.1007/s40894-021-00160-7

@@

Douglas, K., Sutton, R. M., Van Lissa, C. J., Wolfgang, S., Jannis, K., Maximilian, A., ... & N Pontus, L. (2023). Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing. EJSP. DOI: 10.1002/ejsp.2968

>>>>

**Theory**: A model that describes the nature of relations between several phenomena in sufficient detail that it allows for the derivation of quantifiable hypotheses that can be subject to (severe) testing.

* Most social scientific theories are:
    + insufficiently precise
    + too flexible
* Compare to @walasekEvolutionaryModelSensitive2022

@@

* Useful as an instrument of cumulative knowledge acquisition
* Communicating social scientific research to its consumers

@@

## Machine learning vs unguided exploration

```{r, out.width="100%"}
knitr::include_graphics("Figure Machine learning 4.png")
```

@@

* Machine learning can provide nuance by revealing them

@@

* E.g.: assumption that development follows a latent growth curve (LGC).
* SEM forests [@brandmaierTheoryguidedExplorationStructural2016]
* regularized SEM [@jacobucciRegularizedStructuralEquation2016], (Ernst & Peikert 2022)

@@

* Unsupervised learning methods
    + Group individuals to maximize homogeneity of some outcomes
    + "Which individuals are similar?"
    + Non-parametric (e.g., K-means) and model-based (e.g., LCA)
* Tree-based models
    + Group individuals by predictors to maximize homogeneity of outcomes
    + "Why are these individuals similar?"
* Mixed-effects / multilevel models
    + Often model but rarely explain heterogeneity in within-person effects
    + Regularization (e.g., regularized SEM)

@@

Machine learning...

@@

* Variable importance metrics
    + congruent with theoretical assumptions about important predictors?
    + any theoretically important predictors rank low?
    + any undertheorized factors rank high?

* Marginal associations
    + non-linear effects?
    + high importance but flat marginal association?

Comparing predictive performance of simpler parametric model that represents these 'insights' to machine learning model

## From models to theory 2

Data models -> formal theory [@haslbeckModelingPsychopathologyData2021]

* Abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

## Example: TMSR

* Text Mining Systematic Review
* First step in theory development: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
    + Often done by systematic review
* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
    + Unlimited sample size, transparent, objective, reproducible

6653 papers on **emotion regulation** in **adolescence** [age 10-24]

<https://doi.org/10.1007/s40894-021-00160-7>

## Open science

All code and data available at <https://github.com/cjvanlissa/veni_sysrev>

Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]

## Design

1. Manually construct baseline network from relevant theory and prior reviews
2. Identify which constructs co-occur in papers
    + Use POS-tagging to identify nouns and adjectives
    + Use (Google pagerank-based) TextRank to identify bigrams
    + Use regex dictionary to cluster terms
3. Frequency -> Relevance
4. Co-occurrence in publications -> Potential association
5. Compare co-occurrence networks to baseline network

## Baseline network {.smaller}

```{r}
#| label = "figbaseline",
#| warning = FALSE,
#| eval = TRUE,
#| results = "asis",
#| out.width = "100%",
#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
#| the theory)."
knitr::include_graphics("presentation/baseline_network.png") 
```

(a) Theory (b) narrative reviews; transparent nodes indicate constructs also present in theory.

@@

* What are important predictors of adolescents' ER development?
* What is the nature of the association of the predictors with the trajectories?
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>

@@

## Method

SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]

1. Bootstrap sample
2. On each sample, estimate a SEM-model
3. Consider $k$ candidate predictors
4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
5. Average predictions across bootstrap samples

We used 1000 bootstrap samples, $k = 9$

## Relation to other methods

```{r, out.height="50%"}
knitr::include_graphics("groups.png")
```

* Predictors of latent growth model?
  + Only linear differences, few predictors (power)
* Multigroup latent growth model?
  + Non-linear differences, but only one moderator
* Latent class growth model with auxilliary variables?
  + Groups with regard to trajectory, not predictors
* SEM forest
  + Groups with regard to trajectory, based on predictors

@@

## Limitations

1. Problem of induction
2. Causal inference

## Induction {.smaller}

It is impossible to logically induce general principles from specific data

* You may have unobserved counterexamples
* Multiple principles could be consistent with the same data

But it is possible to reason about general principles based on observed patterns in data data

* Popper:
    + "there is no such thing as a logical method of having new ideas"
    + "my arguments in this book are quite independent of [the problem of having new ideas]"
* Machine learning is *one possible method* for generating ideas inductively

## Causal Inference {.smaller}

Machine learning finds patterns in data, irrespective of causality

* If you goal is prediction, this is fine
* But interpreting those patterns assumes causality

Be particularly mindful of:

* Conditioning on colliders
* Multicolinearity

See Pearl 2019

@@

* Machine learning reveals blind spots in existing theory
    + Missing constructs, relative importance, functional form
* ML can detect patterns in published research and primary data
* Does not solve the "inductive problem"
    + Still your job to consolidate insights into theory
* Ignoring causality limits interpretability of machine learning
    + Incorrect direction of causality: meaningless results
    + Conditioning on a collider: biased results

@@

Van Lissa, C. J. (2022). Developmental data science: How machine learning can advance theory formation in Developmental Psychology. Infant and Child Development. DOI:10.1002/icd.2370

@@

Van Lissa, C. J. (2021). Mapping Phenomena Relevant to Adolescent Emotion Regulation: A Text-Mining Systematic Review. Adolescent Research Review. DOI:10.1007/s40894-021-00160-7

@@

Douglas, K., Sutton, R. M., Caspar, V. L., Wolfgang, S., Jannis, K., Maximilian, A., ... & N Pontus, L. (2023). Identifying Important Individual-and Country-Level Predictors of Conspiracy Theorizing: A Machine Learning Analysis. European Journal of Social Psychology. DOI: 10.1002/ejsp.2968
